---
title: "Final Exam"
author: "Konstantin Krönauer"
date: "23-June-2024"
output: html_notebook
---

# Introduction

Chronic Kidney Disease (CKD) is a critical public health issue that affects millions of people worldwide. Early detection and management of CKD can significantly improve patient outcomes. This project aims to develop a predictive model to classify patients as having CKD or not based on various clinical features.

# Data Description

The dataset used for this project contains clinical features and laboratory results of patients. The primary objective is to classify patients into two categories: CKD and not CKD.

## Data Preprocessing

Before training the model, several preprocessing steps were performed:
- Handling missing values by removing rows with missing data.
- Converting all features to factors to facilitate modeling with a decision tree.
- Removing empty and rare classes to ensure the model's robustness.

```{r}
  if (!require(rpart.plot)) install.packages("rpart.plot", dependencies=TRUE)
  if (!require(caret)) install.packages("caret", dependencies=TRUE)
  if (!require(tidyverse)) install.packages("tidyverse", dependencies=TRUE)
  
  library(caret)
  library(ggplot2)
  library(stringr)
  library(tidyverse)
  library(rpart)
  library(rpart.plot)
  library(e1071)


# Read the dataset
data <- read.csv("ckd-dataset.csv")

# Display the first few rows of the dataset
head(data)

# Remove rows with missing values
data <- na.omit(data)

# Convert all features to factors
data <- data %>% mutate(across(everything(), as.factor))

# Remove empty and rare classes
data <- data[!data$class %in% c("", "discrete"), ]

# Verify the structure of the dataset
str(data)

```

# Model Training

Decision tree model to predict CKD status. The dataset was split into training (80%) and testing (20%) sets

```{r}
#Split the data into training and testing sets
    
set.seed(123)
trainIndex <- createDataPartition(data$class, p = .8, list = FALSE, times = 1)

dataTrain <- data[trainIndex, ]
dataTest <- data[-trainIndex, ]

#Train the model
model <- rpart(class ~ ., data = dataTrain, method = "class")

# Plot the decision tree
rpart.plot(model)

```

# Model Evaluation
The model's performance was evaluated using the test set. The confusion matrix and accuracy metrics were used to assess the model's performance.

```{r}

# Make predictions
predictions <- predict(model, dataTest, type = "class")
predictions <- droplevels(predictions)
dataTest$class <- factor(dataTest$class, levels = c("ckd", "notckd"))

# Confusion matrix
confMatrix <- confusionMatrix(predictions, dataTest$class)
print(confMatrix)

# Calculate accuracy
accuracy <- sum(predictions == dataTest$class) / nrow(dataTest)
print(paste("Accuracy:", round(accuracy, 4)))


```

The model achieved an accuracy of 100%, indicating perfect classification on the test set. The confusion matrix further confirmed this with all instances correctly classified.

# Feature Importance
The importance of each feature in predicting CKD was analyzed. The top features contributing to the model's predictions were affected, hemo, pcv, stage, grf, and rbcc.

```{r}


var_importance <- varImp(model)
print(var_importance)
plot(var_importance)

# Variable importance directly from rpart model
importance <- model$variable.importance
print(importance)
barplot(importance, main = "Variable Importance", col = "lightblue", las = 2)


```

Interpretation of Findings
The decision tree model indicates that the most critical factors for predicting CKD are:

affected: Overall health status.
hemo: Hemoglobin levels.
pcv: Packed cell volume.
stage: Stage of CKD.
grf: Glomerular filtration rate.
rbcc: Red blood cell count.
These findings align with medical knowledge, where hemoglobin levels, packed cell volume, and glomerular filtration rate are significant indicators of kidney function.

Conclusion
The decision tree model developed in this project accurately classifies patients with CKD based on clinical features. The model's high accuracy suggests it can be a valuable tool for early CKD detection. Future work could involve validating the model on a larger, more diverse dataset and exploring other machine learning algorithms for potentially improved performance.

This analysis underscores the importance of regular monitoring of kidney function and related clinical parameters in at-risk populations.
